%% Copyright (c) 2002, 2010 Sam Williams
%% Copyright (c) 2010 Richard M. Stallman
%% Permission is granted to copy, distribute and/or modify this
%% document under the terms of the GNU Free Documentation License,
%% Version 1.3 or any later version published by the Free Software
%% Foundation; with no Invariant Sections, no Front-Cover Texts, and
%% no Back-Cover Texts. A copy of the license is included in the
%% file called ``gfdl.tex''.


\chapter{Open Source} \label{chapter:open source}

[RMS: In this chapter only, I have deleted some quotations.  The
material deleted was about open source and didn't relate to my life or
my work.]

In November, 1995, Peter Salus, a member of the Free Software Foundation and author of the 1994 book, \textit{A Quarter Century of Unix}, issued a call for papers to members of the GNU Project's ``system-discuss'' mailing list. Salus, the conference's scheduled chairman, wanted to tip off fellow hackers about the upcoming Conference on Freely Redistributable Software in Cambridge, Massachusetts. Slated for February, 1996, and sponsored by the Free Software Foundation, the event promised to be the first engineering conference solely dedicated to free software and, in a show of unity with other free software programmers, welcomed papers on ``any aspect of GNU, Linux, NetBSD, 386BSD, FreeBSD, Perl, Tcl/tk, and other tools for which the code is accessible and redistributable.'' Salus wrote:

\begin{quote}
Over the past 15 years, free and low-cost software has become ubiquitous. This conference will bring together implementers of several different types of freely redistributable software and publishers of such software (on various media). There will be tutorials and refereed papers, as well as keynotes by Linus Torvalds and Richard Stallman.\endnote{See Peter Salus, ``FYI-Conference on Freely Redistributable Software, 2/2, Cambridge'' (1995) (archived by Terry Winograd), \url{http://bat8.inria.fr/~lang/hotlist/free/licence/fsf96/call-for-papers.html}.}
\end{quote}

Among the recipients of Salus' email was conference committee member Eric S. Raymond. Although not the leader of a project or company like the various other members of the list, Raymond had built a tidy reputation within the hacker community for some software projects and as editor of \textit{The New Hacker's Dictionary}, a greatly enlarged version of \textit{The Hacker's Dictionary} published a decade earlier by Guy Steele.

For Raymond, the 1996 conference was a welcome event. Although he did not thoroughly support the free software movement's ideas, he had contributed to some GNU programs, in particular to GNU Emacs. Those contributions stopped in 1992, when Raymond demanded authority to make changes in the official GNU version of GNU Emacs without discussing them with Stallman, who was directly in charge of Emacs development.  Stallman rejected the demand, and Raymond accused Stallman of ``micro-management.''  ``Richard kicked up a fuss about my making unauthorized modifications when I was cleaning up the Emacs LISP libraries,'' Raymond recalls. ``It frustrated me so much that I decided I didn't want to work with him anymore.''

Despite the falling out, Raymond remained active in the free software community. So much so that when Salus suggested a conference pairing Stallman and Torvalds as keynote speakers, Raymond eagerly seconded the idea. With Stallman representing the older, wiser contingent of ITS/Unix hackers and Torvalds representing the younger, more energetic crop of Linux hackers, the pairing indicated a symbolic show of unity that could only be beneficial, especially to ambitious younger (i.e., below 40) hackers such as Raymond. ``I sort of had a foot in both camps,'' Raymond says.

By the time of the conference, the tension between those two camps had become palpable. Both groups had one thing in common, though: the conference was their first chance to meet the Finnish \textit{wunderkind} in the flesh. Surprisingly, Torvalds proved himself to be a charming, affable speaker. Possessing only a slight Swedish accent, Torvalds surprised audience members with his quick, self-effacing wit.\endnote{Although Linus Torvalds is Finnish, his mother tongue is Swedish. ``The Rampantly Unofficial Linus FAQ'' at \url{http://catb.org/~esr/faqs/linus/} offers a brief explanation:

\begin{quote}
Finland has a significant (about 6\%) Swedish-speaking minority population. They call themselves \textit{finlandssvensk} or \textit{finlandssvenskar} and consider themselves Finns; many of their families have lived in Finland for centuries. Swedish is one of Finland's two official languages.
\end{quote}} 

Even more surprising, says Raymond, was Torvalds' equal willingness to take potshots at other prominent hackers, including the most prominent hacker of all, Richard Stallman. By the end of the conference, Torvalds' half-hacker, half-slacker manner was winning over older and younger conference-goers alike.

``It was a pivotal moment,'' recalls Raymond. ``Before 1996, Richard was the only credible claimant to being the ideological leader of the entire culture. People who dissented didn't do so in public. The person who broke that taboo was Torvalds.''

The ultimate breach of taboo would come near the end of the show. During a discussion on the growing market dominance of Microsoft Windows or some similar topic, Torvalds admitted to being a fan of Microsoft's PowerPoint slideshow software program. From the perspective of old-line software purists, it was like bragging about one's slaves at an abolitionist conference. From the perspective of Torvalds and his growing band of followers, it was simply common sense. Why shun convenient proprietary software programs just to make a point? They didn't agree with the point anyway. When freedom requires a sacrifice, those who don't care about freedom see the sacrifice as self-denial, rather than as a way to obtain something important.  Being a hacker wasn't about self-denial, it was about getting the job done, and ``the job,'' for them, was defined in practical terms.

``That was a pretty shocking thing to say,'' Raymond remembers. ``Then again, he was able to do that, because by 1995 and 1996, he was rapidly acquiring clout.''

Stallman, for his part, doesn't remember any tension at the 1996 conference; he probably wasn't present when Torvalds made that statement.  But he does remember later feeling the sting of Torvalds' celebrated ``cheekiness.'' ``There was a thing in the Linux documentation which says print out the GNU coding standards and then tear them up,'' says Stallman, recalling one example. ``When you look closely, what he disagreed with was the least important part of it, the recommendation for how to indent C code.''

``OK, so he disagrees with some of our conventions. That's fine, but he picked a singularly nasty way of saying so. He could have just said, `Here's the way I think you should indent your code.' Fine. There should be no hostility there.''

For Raymond, the warm reception other hackers gave to Torvalds' comments confirmed a suspicion: the dividing line separating Linux developers from GNU developers was largely generational. Many Linux hackers, like Torvalds, had grown up in a world of proprietary software. They had begun contributing to free software without perceiving any injustice in nonfree software.  For most of them, nothing was at stake beyond convenience.  Unless a program was technically inferior, they saw little reason to reject it on licensing issues alone. Some day hackers might develop a free software alternative to PowerPoint. Until then, why criticize PowerPoint or Microsoft; why not use it?

This was an example of the growing dispute, within the free software community, between those who valued freedom as such, and those who mainly valued powerful, reliable software.  Stallman referred to the two camps as political parties within the community, calling the former the ``freedom party.''  The supporters of the other camp did not try to name it, so Stallman disparagingly called it the ``bandwagon party'' or the ``success party,'' because many of them presented ``more
users'' as the primary goal.

In the decade since launching the GNU Project, Stallman had built up a fearsome reputation as a programmer. He had also built up a reputation for intransigence both in terms of software design and people management. This was partly true, but the reputation provided a convenient excuse that anyone could cite if Stallman did not do as he wished.  The reputation has been augmented by mistaken guesses.

For example, shortly before the 1996 conference, the Free Software Foundation experienced a full-scale staff defection.  Brian Youmans, a current FSF staffer hired by Salus in the wake of the resignations, recalls the scene: ``At one point, Peter [Salus] was the only staff member working in the office.''  The previous staff were unhappy with the executive director; as Bryt Bradley told her friends in December, 1995:

\begin{quote}
[name omitted] (the Executive Director of the FSF) decided to come back from Medical/Political Leave last week. The office staff (Gena Bean, Mike Drain, and myself) decided we could not work with her as our supervisor because of the many mistakes she had made in her job tasks prior to her taking a leave.  Also, there had been numerous instances where individuals were threatened with inappropriate firing and there  were many instances of what we felt were verbal abuse from her to ALL members of the office staff.  We requested (many times) that she not come back as our supervisor, but stated that we were willing to work with her as a co-worker.  Our requests were ignored.  We quit.
\end{quote}

The executive director in question then gave Stallman an ultimatum: give her total autonomy in the office or she would quit.  Stallman, as president of the FSF, declined to give her total control over its activities, so she resigned, and he recruited in Peter Salus to replace her.

When Raymond, an outsider, learned that these people had left the FSF, he presumed Stallman was at fault.  This provided confirmation for his theory that Stallman's personality was the cause of any and all problems in the GNU Project.

Raymond had another theory: recent delays such as the Hurd and recent troubles such as the Lucid-Emacs schism reflected problems normally associated with software project management, not software code development.

Shortly after the Freely Redistributable Software Conference, Raymond began working on his own pet software project, a mail utility called ``fetchmail.'' Taking a cue from Torvalds, Raymond issued his program with a tacked-on promise to update the source code as early and as often as possible. When users began sending in bug reports and feature suggestions, Raymond, at first anticipating a tangled mess, found the resulting software surprisingly sturdy. Analyzing the success of the Torvalds approach, Raymond issued a quick analysis: using the Internet as his ``petri dish'' and the harsh scrutiny of the hacker community as a form of natural selection, Torvalds had created an evolutionary model free of central planning.

What's more, Raymond decided, Torvalds had found a way around Brooks' Law. First articulated by Fred P. Brooks, manager of IBM's OS/360 project and author of the 1975 book, \textit{The Mythical Man-Month}, Brooks' Law held that adding developers to a project only resulted in further project delays. Believing as most hackers that software, like soup, benefits from a limited number of cooks, Raymond sensed something revolutionary at work. In inviting more and more cooks into the kitchen, Torvalds had actually found a way to make the resulting software \textit{better}.\endnote{Brooks' Law is the shorthand summary of the following quote taken from Brooks' book:

\begin{quote}
Since software construction is inherently a systems effort -- an exercise in complex interrelationships -- communication effort is great, and it quickly dominates the decrease in individual task time brought about by partitioning. Adding more men then lengthens, not shortens, the schedule.
\end{quote}

\noindent See Fred P. Brooks, \textit{The Mythical Man-Month} (Addison Wesley Publishing, 1995).}

Raymond put his observations on paper. He crafted them into a speech, which he promptly delivered before a group of friends and neighbors in Chester County, Pennsylvania. Dubbed ``The Cathedral and the Bazaar,'' the speech contrasted the ``Bazaar'' style originated by Torvalds with the ``Cathedral'' style generally used by everyone else.

Raymond says the response was enthusiastic, but not nearly as enthusiastic as the one he received during the 1997 Linux Kongress, a gathering of GNU/Linux users in Germany the next spring.

``At the Kongress, they gave me a standing ovation at the end of the speech,'' Raymond recalls. ``I took that as significant for two reasons. For one thing, it meant they were excited by what they were hearing. For another thing, it meant they were excited even after hearing the speech delivered through a language barrier.''

Eventually, Raymond would convert the speech into a paper, also titled ``The Cathedral and the Bazaar.'' The paper drew its name from Raymond's central analogy. Previously, programs were ``cathedrals,'' impressive, centrally planned monuments built to stand the test of time. Linux, on the other hand, was more like ``a great babbling bazaar,'' a software program developed through the loose decentralizing dynamics of the Internet.

Raymond's paper associated the Cathedral style, which he and Stallman and many others had used, specifically with the GNU Project and Stallman, thus casting the contrast between development models as a comparison between Stallman and Torvalds. Where Stallman was his chosen example of the classic cathedral architect -- i.e., a programming ``wizard'' who could disappear for 18 months and return with something like the GNU C Compiler -- Torvalds was more like a genial dinner-party host. In letting others lead the Linux design discussion and stepping in only when the entire table needed a referee, Torvalds had created a development model very much reflective of his own laid-back personality. From Torvalds' perspective, the most important managerial task was not imposing control but keeping the ideas flowing.

Summarized Raymond, ``I think Linus's cleverest and most consequential hack was not the construction of the Linux kernel itself, but rather his invention of the Linux development model.''\endnote{See Eric Raymond, ``The Cathedral and the Bazaar'' (1997).}

If the paper's description of these two styles of development was perceptive, its association of the Cathedral model specifically with Stallman (rather than all the others who had used it, including Raymond himself) was sheer calumny.  In fact, the developers of some GNU packages including the GNU Hurd had read about and adopted Torvalds' methods before Raymond tried them, though without analyzing them further and publicly championing them as Raymond's paper did. Thousands of hackers, reading Raymond's article, must have been led to a negative attitude towards GNU by this smear.

In summarizing the secrets of Torvalds' managerial success, Raymond attracted the attention of other members of the free software community for whom freedom was not a priority.  They sought to interest business in the use and development of free software, and to do so, decided to cast the issue in terms of the values that appeal to business: powerful, reliable, cheap, advanced.  Raymond became the best-known proponent of these ideas, and they reached the management of Netscape, whose proprietary browser was losing market share to Microsoft's equally proprietary Internet Explorer.  Intrigued by a speech by Raymond, Netscape executives took the message back to corporate headquarters. A few months later, in January, 1998, the company announced its plan to publish the source code of its flagship Navigator web browser in the hopes of enlisting hacker support in future development.

When Netscape CEO Jim Barksdale cited Raymond's ``Cathedral and the Bazaar'' essay as a major influence upon the company's decision, the company instantly elevated Raymond to the level of hacker celebrity. He invited a few people including Larry Augustin, founder of VA Research which sold workstations with the GNU/Linux operating system preinstalled; Tim O'Reilly,  founder of the publisher O'Reilly \& Associates; and Christine Peterson, president of the Foresight Institute, a Silicon Valley think tank specializing in nanotechnology, to talk. ``The meeting's agenda boiled down to one item: how to take advantage of Netscape's decision so that other companies might follow suit?''

Raymond doesn't recall the conversation that took place, but he does remember the first complaint addressed. Despite the best efforts of Stallman and other hackers to remind people that the word ``free'' in free software stood for freedom and not price, the message still wasn't getting through. Most business executives, upon hearing the term for the first time, interpreted the word as synonymous with ``zero cost,'' tuning out any follow-up messages in short order. Until hackers found a way to get past this misunderstanding, the free software movement faced an uphill climb, even after Netscape.

Peterson, whose organization had taken an active interest in advancing the free software cause, offered an alternative: ``open source.''

Looking back, Peterson says she came up with the ``open source'' term while discussing Netscape's decision with a friend in the public relations industry. She doesn't remember where she came upon the term or if she borrowed it from another field, but she does remember her friend disliking the term.\endnote{See Malcolm Maclachlan, ``Profit Motive Splits Open Source Movement,'' \textit{TechWeb News} (August 26, 1998), \url{http://www.techweb.com/article/showArticle?articleID=29102344}.}

At the meeting, Peterson says, the response was dramatically different. ``I was hesitant about suggesting it,'' Peterson recalls. ``I had no standing with the group, so started using it casually, not highlighting it as a new term.'' To Peterson's surprise, the term caught on. By the end of the meeting, most of the attendees, including Raymond, seemed pleased by it.

Raymond says he didn't publicly use the term ``open source'' as a substitute for ``free software'' until a day or two after the Mozilla launch party, when O'Reilly had scheduled a meeting to talk about free software. Calling his meeting ``the Freeware Summit,'' O'Reilly says he wanted to direct media and community attention to the other deserving projects that had also encouraged Netscape to release Mozilla. ``All these guys had so much in common, and I was surprised they didn't all know each other,'' says O'Reilly. ``I also wanted to let the world know just how great an impact the free software culture had already made. People were missing out on a large part of the free software tradition.''

In putting together the invite list, however, O'Reilly made a decision that would have long-term political consequences. He decided to limit the list to west-coast developers such as Wall, Eric Allman, creator of sendmail, and Paul Vixie, creator of BIND. There were exceptions, of course: Pennsylvania-resident Raymond, who was already in town thanks to the Mozilla launch, earned an quick invite. So did Virginia-resident Guido van Rossum, creator of Python. ``Frank Willison, my editor in chief and champion of Python within the company, invited him without first checking in with me,'' O'Reilly recalls. ``I was happy to have him there, but when I started, it really was just a local gathering.''

For some observers, the unwillingness to include Stallman's name on the list qualified as a snub. ``I decided not to go to the event because of it,'' says Perens, remembering the summit. Raymond, who did go, says he argued for Stallman's inclusion to no avail. The snub rumor gained additional strength from the fact that O'Reilly, the event's host, had feuded publicly with Stallman over the issue of software-manual copyrights. Prior to the meeting, Stallman had argued that free software manuals should be as freely copyable and modifiable as free software programs. O'Reilly, meanwhile, argued that a value-added market for nonfree books increased the utility of free software by making it more accessible to a wider community. The two had also disputed the title of the event, with Stallman insisting on ``Free Software'' rather than ``Freeware.''  The latter term most often refers to programs which are available gratis, but which are not free software because their source code is not released.

Looking back, O'Reilly doesn't see the decision to leave Stallman's name off the invite list as a snub. ``At that time, I had never met Richard in person, but in our email interactions, he'd been inflexible and unwilling to engage in dialogue. I wanted to make sure the GNU tradition was represented at the meeting, so I invited John Gilmore and Michael Tiemann, whom I knew personally, and whom I knew were passionate about the value of the GPL but seemed more willing to engage in a frank back-and-forth about the strengths and weaknesses of the various free software projects and traditions. Given all the later brouhaha, I do wish I'd invited Richard as well, but I certainly don't think that my failure to do so should be interpreted as a lack of respect for the GNU Project or for Richard personally.''

Snub or no snub, both O'Reilly and Raymond say the term ``open source'' won over just enough summit-goers to qualify as a success. The attendees shared ideas and experiences and brainstormed on how to improve free software's image. Of key concern was how to point out the successes of free software, particularly in the realm of Internet infrastructure, as opposed to playing up the GNU/Linux challenge to Microsoft Windows. But like the earlier meeting at VA, the discussion soon turned to the problems associated with the term ``free software.'' O'Reilly, the summit host, remembers a comment from Torvalds, a summit attendee.

``Linus had just moved to Silicon Valley at that point, and he explained how only recently that he had learned that the word `free' had two meanings -- free as in `libre' and free as in `gratis' -- in English.''

Michael Tiemann, founder of Cygnus, proposed an alternative to the troublesome ``free software'' term: sourceware. ``Nobody got too excited about it,'' O'Reilly recalls. ``That's when Eric threw out the term `open source.'\hspace{0.01in}''

Although the term appealed to some, support for a change in official terminology was far from unanimous. At the end of the one-day conference, attendees put the three terms -- free software, open source, or sourceware -- to a vote. According to O'Reilly, 9 out of the 15 attendees voted for ``open source.'' Although some still quibbled with the term, all attendees agreed to use it in future discussions with the press. ``We wanted to go out with a solidarity message,'' O'Reilly says.

The term didn't take long to enter the national lexicon. Shortly after the summit, O'Reilly shepherded summit attendees to a press conference attended by reporters from the \textit{New York Times}, the \textit{Wall Street Journal}, and other prominent publications. Within a few months, Torvalds' face was appearing on the cover of \textit{Forbes} magazine, with the faces of Stallman, Perl creator Larry Wall, and Apache team leader Brian Behlendorf featured in the interior spread. Open source was open for business.

For summit attendees such as Tiemann, the solidarity message was the most important thing. Although his company had achieved a fair amount of success selling free software tools and services, he sensed the difficulty other programmers and entrepreneurs faced.

``There's no question that the use of the word free was confusing in a lot of situations,'' Tiemann says. ``Open source positioned itself as being business friendly and business sensible. Free software positioned itself as morally righteous. For better or worse we figured it was more advantageous to align with the open source crowd.

Raymond called Stallman after the meeting to tell him about the new term ``open source'' and ask if he would use it.  Raymond says Stallman briefly considered adopting the term, only to discard it. ``I know because I had direct personal conversations about it,'' Raymond says.

Stallman's immediate response was, ``I'll have to think about it.''  The following day he had concluded that the values of Raymond and O'Reilly would surely dominate the future discourse of ``open source,'' and that the best way to keep the ideas of the free software movement in public view was to stick to its traditional term.

Later in 1998, Stallman announced his position: ``open source,'' while helpful in communicating the technical advantages of free software also encouraged speakers to soft-pedal the issue of software freedom. It avoided the unintended meaning of ``gratis software'' and the intended meaning of ``freedom-respecting software'' equally.  As a means for conveying the latter meaning, it was therefore no use.  In effect, Raymond and O'Reilly had given a name to the nonidealistic political party in the community, the one Stallman did not agree with.

In addition, Stallman thought that the ideas of ``open source'' led people to put too much emphasis on winning the support of business.  While such support in itself wasn't necessarily bad in itself, he expected that being too desperate for it would lead to harmful compromises.  ``Negotiation 101 would teach you that if you are desperate to get someone's agreement, you are asking for a bad deal,'' he says.  ``You need to be prepared to say no.''  Summing up his position at the 1999 LinuxWorld Convention and Expo, an event billed by Torvalds himself as a ``coming out party'' for the ``Linux'' community, Stallman implored his fellow hackers to resist the lure of easy compromise.

``Because we've shown how much we can do, we don't have to be desperate to work with companies or compromise our goals,'' Stallman said during a panel discussion. ``Let them offer and we'll accept. We don't have to change what we're doing to get them to help us. You can take a single step towards a goal, then another and then more and more and you'll actually reach your goal. Or, you can take a half measure that means you don't ever take another step, and you'll never get there.''

Even before the LinuxWorld show, however, Stallman was showing an increased willingness to alienate open source supporters. A few months after the Freeware Summit, O'Reilly hosted its second annual Perl Conference. This time around, Stallman was in attendance. During a panel discussion lauding IBM's decision to employ the free software Apache web server in its commercial offerings, Stallman, taking advantage of an audience microphone, made a sharp denunciation of panelist John Ousterhout, creator of the Tcl scripting language. Stallman branded Ousterhout a ``parasite'' on the free software community for marketing a proprietary version of Tcl via Ousterhout's startup company, Scriptics.  Ousterhout had stated that Scriptics would contribute only the barest minimum of its improvements to the free version of Tcl, meaning it would in effect use that small contribution to win community approval for much a larger amount of nonfree software development.  Stallman rejected this position and denounced Scriptics' plans. ``I don't think Scriptics is necessary for the continued existence of Tcl,'' Stallman said to hisses from the fellow audience members.\endnote{\textit{Ibid.}}

``It was a pretty ugly scene,'' recalls Prime Time Freeware's Rich Morin. ``John's done some pretty respectable things: Tcl, Tk, Sprite. He's a real contributor.''  Despite his sympathies for Stallman and Stallman's position, Morin felt empathy for those troubled by Stallman's discordant words.

Stallman will not apologize.  ``Criticizing proprietary software isn't ugly -- proprietary software is ugly.  Ousterhout had indeed made real contributions in the past, but the point is that Scriptics was going to be nearly 100\% a proprietary software company.  In that conference, standing up for freedom meant disagreeing with nearly everyone. Speaking from the audience, I could only say a few sentences.  The only way to raise the issue so it would not be immediately forgotten was to put it in strong terms.''

``If people rebuke me for `making a scene' when I state a serious criticism of someone's conduct, while calling Torvalds `cheeky' for saying nastier things about trivial matters, that seems like a double standard to me.''

Stallman's controversial criticism of Ousterhout momentarily alienated a potential sympathizer, Bruce Perens. In 1998, Eric Raymond proposed launching the Open Source Initiative, or OSI, an organization that would police the use of the term ``open source'' and provide a definition for companies interested in making their own programs. Raymond recruited Perens to draft the definition.\endnote{See Bruce Perens et al., ``The Open Source Definition,'' The Open Source Initiative (1998), \url{http://www.opensource.org/docs/definition.html}.}

Perens would later resign from the OSI, expressing regret that the organization had set itself up in opposition to Stallman and the FSF. Still, looking back on the need for a free software definition outside the Free Software Foundation's auspices, Perens understands why other hackers might still feel the need for distance. ``I really like and admire Richard,'' says Perens. ``I do think Richard would do his job better if Richard had more balance. That includes going away from free software for a couple of months.''

Stallman's energies would do little to counteract the public-relations momentum of open source proponents. In August of 1998, when chip-maker Intel purchased a stake in GNU/Linux vendor Red Hat, an accompanying \textit{New York Times} article described the company as the product of a movement ``known alternatively as free software and open source.''\endnote{See Amy Harmon, ``For Sale: Free Operating System,'' \textit{New York Times} (September 28, 1998), \url{http://www.nytimes.com/library/tech/98/09/biztech/articles/28linux.html}.} Six months later, a John Markoff article on Apple Computer was proclaiming the company's adoption of the ``open source'' Apache server in the article headline.\endnote{See John Markoff, ``Apple Adopts `Open Source' for its Server Computers,'' \textit{New York Times} (March 17, 1999), \url{http://www.nytimes.com/library/tech/99/03/biztech/articles/17apple.html}.}

Such momentum would coincide with the growing momentum of companies that actively embraced the ``open source'' term. By August of 1999, Red Hat, a company that now eagerly billed itself as ``open source,'' was selling shares on Nasdaq. In December, VA Linux -- formerly VA Research -- was floating its own IPO to historic effect. Opening at \$30 per share, the company's stock price exploded past the \$300 mark in initial trading only to settle back down to the \$239 level. Shareholders lucky enough to get in at the bottom and stay until the end experienced a 698\% increase in paper wealth, a Nasdaq record.  Eric Raymond, as a board member, owned shares worth \$36 million.  However, these high prices were temporary; they tumbled when the dot-com boom ended.

The open source proponents' message was simple: all you need, to sell the free software concept, is to make it business-friendly.  They saw Stallman and the free software movement as fighting the market; they sought instead to leverage it. Instead of playing the role of high-school outcasts, they had played the game of celebrity, magnifying their power in the process.

These methods won great success for open source, but not for the ideals of free software.  What they had done to ``spread the message'' was to omit the most important part of it: the idea of freedom as an ethical issue.  The effects of this omission are visible today: as of 2009, nearly all GNU/Linux distributions include proprietary programs, Torvalds' version of Linux contains proprietary firmware programs, and the company formerly called VA Linux bases its business on proprietary software.  Over half of all the world's web servers run some version of Apache, and the usual version of Apache is free software, but many of those sites run a proprietary modified version distributed by IBM.

``On his worst days Richard believes that Linus Torvalds and I conspired to hijack his revolution,'' Raymond says. ``Richard's rejection of the term open source and his deliberate creation of an ideological fissure in my view comes from an odd mix of idealism and territoriality. There are people out there who think it's all Richard's personal ego. I don't believe that. It's more that he so personally associates himself with the free software idea that he sees any threat to that as a threat to himself.''

Stallman responds, ``Raymond misrepresents my views: I don't think Torvalds `conspired' with anyone, since being sneaky is not his way. However, Raymond's nasty conduct is visible in those statements themselves.   Rather than respond to my views (even as he claims they are) on their merits, he proposes psychological interpretations for them. He attributes the harshest interpretation to unnamed others, then `defends' me by proposing a slightly less derogatory one.  He has often `defended' me this way.''

Ironically, the success of open source and open source advocates such as Raymond would not diminish Stallman's role as a leader -- but it would lead many to misunderstand what he is a leader of.  Since the free software movement lacks the corporate and media recognition of open source, most users of GNU/Linux do not hear that it exists, let alone what its views are.  They have heard the ideas and values of open source, and they never imagine that Stallman might have different ones.  Thus he receives messages thanking him for his advocacy of ``open source,'' and explains in response that he has never been a supporter of that, using the occasion to inform the sender about free software.

Some writers recognize the term ``free software'' by using the term ``FLOSS,'' which stands for ``Free/Libre and Open Source Software.''  However, they often say there is a single ``FLOSS'' movement, which is like saying that the U.S. has a ``Liberal/Conservative'' movement, and the views they usually associate with this supposed single movement are the open source views they have heard.

Despite all these obstacles, the free software movement does make its ideas heard sometimes, and continues to grow in absolute terms.  By sticking to its guns, and presenting its ideas in contrast to those of open source, it gains ground. ``One of Stallman's primary character traits is the fact he doesn't budge,'' says Ian Murdock. ``He'll wait up to a decade for people to come around to his point of view if that's what it takes.''

Murdock, for one, finds that unbudgeable nature both refreshing and valuable. Stallman may no longer be the solitary leader of the free software movement, but he is still the polestar of the free software community. ``You always know that he's going to be consistent in his views,'' Murdock says. ``Most people aren't like that. Whether you agree with him or not, you really have to respect that.''

\theendnotes
\setcounter{endnote}{0}
